flowCore: a Bioconductor package for high throughput flow cytometry

### 2009-03-23 paper/bmc_article.tex
reviewer recommended updates
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@38132 bc3139a8-67e5-0310-9ffc-ced21a209358

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
  \paragraph*{Background:} Recent advances in automation technologies
  have enabled the use of flow cytometry high content screening
  (FH-HCS), generating large complex data sets often in clinical
  trials or drug discovery settings. However, data management and data
  analysis methods have not advanced sufficiently far from the initial
  small-scale studies to support modeling in the presence of multiple
  covariates.

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Automation technologies developed during the last several years have
enabled the use of flow cytometry (FCM) to generate large, complex
data sets in both basic and clinical research applications
\cite{brinkman2007hcf}. A serious bottleneck in the interpretation of
existing studies and the application of flow cytometry high content
screening (FC-HCS) to even larger, more complex problems is that data
management and data analysis methods have not advanced sufficiently
far from the methods developed for applications of FCM
 to small-scale,
tube-based studies \cite{mahnke2007omi}. In particular, the data often
need to be organized into groups of samples based on combinations of
additional covariates and similar operations need to be applied to
these groups in a transparent and reproducible manner. Furthermore,
the growing depth of knowledge in the field of immunology, for
instance the characterization of distinct human T-cell sub-population
\cite{Appay}, clearly argues for more systematic approaches.
%% add one of the last roederer's paper that "request" algo and
%% automation for FC-HCS data analysis.

@@

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures.
In addition, functions or methods that work on those classes allow for
interaction and manipulation. Many of the currently available software
solutions offer only limited support for such self-contained
structures, or make use of containers like XML or binary storage
formats that are designed specifically for particular user interfaces
and hence not easily amenable to programmatical access. In addition,
the closed-source nature of these products often makes them
impractical to integrate into analysis pipelines. In this manuscript
we describe classes for FCM data analysis and their implementation in
R, however, they could just as easily be implemented in any other
language (e.g., Java, C++). Software written in those languages could
use similar data structures, thereby simplifying communication and the
interchange of data between analysis tools.

@@

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~1\ref{fig1:FrameWork}). Moreover, it is
a software implementation of the Gating-Markup Language Candidate
Recommendation (Gating-ML CR), an emerging standard developed in collaboration
with the International Society for Analytical Cytology (ISAC) Data
Standards Task Force, which makes it possible to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Data transformations}
    \begin{table}[ht]
      \caption{\label{table1} Data transformations implemented in \Rpackage{flowCore}. Within these formulas, $x$ is the variable
        corresponding to value being transformed, $y$ is the result of the
        transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
        constants affecting the transformation function, $e$ is the base of
        the natural logarithm, and $root$ is a root-finding algorithm. Other transformations can easily be implemented in R.}
      \begin{center}
        \begin{tabular}{|l|l|}
          \hline
          \multicolumn{2}{|c|}{Data Transformations} \\
          \hline
          linear & $ax + b$ \\
          quadratic & $ax^2 + bx + c$ \\
          natural logarithm & $log_e(x)(r/d)$ \\
          logarithm & $log_b(x)(r/d)$ \\
          biexponential & $root(ae^{(b*y)}-ce^{(-d+y)}+f)$ \\
          logicle& $root(Te^{-(m-w)}(e^{(y-w)}-p^2e^{-(y-w)/p}+p^2-1)$ \\
          truncate & $x_{x{\leq}a} = a$ \\
          scale & $(x-a)/(b-a)$ \\
          split-scale & combination of linear and logarithm \\
          arcsinh & $arcsinh(a + bx)+c$ \\
          \hline
        \end{tabular}
      \end{center}
    \end{table}

>>>>

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
  \paragraph*{Background:} Recent advances in automation technologies
  have enabled the use of flow cytometry for high throughput
  screening, generating large complex data sets often in clinical
  trials or drug discovery settings. However, data management and data
  analysis methods have not advanced sufficiently far from the initial
  small-scale studies to support modeling in the presence of multiple
  covariates.

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Automation technologies developed during the last several years have
enabled the use of flow cytometry (FCM) to generate large, complex
data sets in both basic and clinical research applications
\cite{brinkman2007hcf}. A serious bottleneck in the interpretation of
existing studies and the application of high throughput FCM to even
larger, more complex problems is that data management and data
analysis methods have not advanced sufficiently far from the methods
developed for applications of FCM to small-scale, tube-based studies
\cite{mahnke2007omi}. In particular, the data often need to be
organized into groups of samples based on combinations of additional
covariates and similar operations need to be applied to these groups
in a transparent and reproducible manner. Furthermore, the growing
depth of knowledge in the field of immunology, for instance the
characterization of distinct human T-cell sub-population \cite{Appay},
clearly argues for more systematic approaches.
%% add one of the last roederer's paper that "request" algo and
%% automation for FC-HCS data analysis.

@@

The notion of classes from an object-oriented programming language
provides one coherent way to describe these richer data structures.
In addition, functions or methods that work on those classes allow for
interaction and manipulation. Many of the currently available software
solutions offer only limited support for such self-contained
structures, or make use of binary storage containers that are designed
specifically for the needs of particular user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of these products often makes them impractical to
integrate into analysis pipelines. In this manuscript we describe
classes for FCM data analysis and their implementation in R, however,
they could just as easily be implemented in any other language (e.g.,
Java, C++). Software written in those languages could use similar data
structures, thereby simplifying communication and the interchange of
data between analysis tools.

@@

The \Rpackage{flowCore} framework presented here can import and
process raw data FCS files along with their complete set of
file-specific metadata (Figure~1\ref{fig1:FrameWork}). Moreover, it is
a software implementation of the Gating Markup Language Candidate
Recommendation, an emerging standard developed in collaboration
with the International Society for Analytical Cytology (ISAC) Data
Standards Task Force, which makes it possible to integrate
\Rpackage{flowCore} in existing work flows and to communicate with any
other FCM tool that adheres to the proposed standard
\cite{SpidlenInPressCytometryA}. Adherence to standards also plays a
critical role in the ability of new methods based on
\Rpackage{flowCore} to find their way back into the standard practices
for FCM data analysis.

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Data transformations}
    \begin{table}[ht]
      \caption{\label{table1} Data transformations implemented in \Rpackage{flowCore}. Within these formulas, $x$ is the variable
        corresponding to value being transformed, $y$ is the result of the
        transformation, $a$, $b$, $c$, $d$, $f$, $p$, $m$, $T$, and $w$, are
        constants affecting the transformation function, $e$ is the base of
        the natural logarithm, and $root$ is a root-finding algorithm. Other transformations can easily be implemented in R.}
      \begin{center}
        \begin{tabular}{|l|l|}
          \hline
          \multicolumn{2}{|c|}{Data Transformations} \\
          \hline
          linear & $ax + b$ \\
          quadratic & $ax^2 + bx + c$ \\
          natural logarithm & $log_e(x)(r/d)$ \\
          logarithm & $log_b(x)(r/d)$ \\
          biexponential & $ae^{(b*x)}-ce^{(-d+x)}+f$ \\
          logicle& $Te^{-(m-w)}(e^{(x-w)}-p^2e^{-(x-w)/p}+p^2-1)$ \\
          truncate & $x_{x{\leq}a} = a$ \\
          scale & $(x-a)/(b-a)$ \\
          arcsinh & $arcsinh(a + bx)+c$ \\
          \hline
        \end{tabular}
      \end{center}
    \end{table}
